The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-07 script/de/chapters/introduction.tex
new float cryptoprocedure for crypto algorithms and procedures (chapter 5); fixed links to un-numbered chapters with phantomsection
CT script: Minor updates in the preface (de+en), chap 1 (de+en) and chap 4 (Sage samples, en)

% --------------------------------------------------------------------------
\newpage
%\addcontentsline{toc}{chapter}{Vorwort zur 8. Auflage des CrypTool-Skripts} % be_2006: So springt er auf S. 3 statt auf S. 9 !!
%\chapter*{Vorwort zur 10. Auflage des CrypTool-Skripts\footnote{Zur 9. Auflage wurde kein extra Vorwort verfasst.}}
\chapter*{Vorwort zur 10. Auflage des CrypTool-Skripts}
\addcontentsline{toc}{chapter}{Vorwort zur 10. Auflage des CrypTool-Skripts}

@@

%In dieser Ausgabe des Skripts wurden einige Themen ergänzt und andere auf
%den aktuellen Stand gebracht (z.B. die Zusammenfassungen zu aktuellen 
%Forschungsergebnissen):                   
In dieser Ausgabe des Skripts wurden wieder einige Themen korrigiert,
ergänzt und auf den aktuellen Stand gebracht, z.B.:
\vspace{-7pt}
\begin{itemize}
  \item die Suche nach den größten Primzahlen,
%  \item Fortschritte bei der Kryptoanalyse von Hashverfahren 
%        (Kap. \ref{collision-attacks-against-sha-1}) und
%  \item Fortschritte bei den Ideen für neue Kryptoverfahren (RSA-Nachfolger)
%	(Kap. \ref{xxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxxxxxxx} und
  \item die Auflistung, in welchen Filmen und Romanen Kryptographie eine
        wesentliche Rolle spielt (siehe Anhang \ref{s:appendix-movies}),
        und Kurioses zu Primzahlen (siehe \ref{HT-Quaint-curious-Primes-usage}).
\end{itemize}

@@

% --------------------------------------------------------------------------
\newpage
%\addcontentsline{toc}{chapter}{Einführung -- Zusammenspiel von Skript
%        und CrypTool}  % be_2006: So springt er auf S. 9 statt auf S. 10 !!
\chapter*{Einführung -- Zusammenspiel von Skript und CrypTool}
\addcontentsline{toc}{chapter}{Einführung -- Zusammenspiel von Skript
                               und CrypTool}

>>>>

% --------------------------------------------------------------------------
\clearpage\phantomsection
\addcontentsline{toc}{chapter}{Vorwort zur 10. Auflage des CrypTool-Skripts}
\chapter*{Vorwort zur 10. Auflage des CrypTool-Skripts}

@@

%In dieser Ausgabe des Skripts wurden einige Themen ergänzt und andere auf
%den aktuellen Stand gebracht (z.B. die Zusammenfassungen zu aktuellen 
%Forschungsergebnissen):                   
In dieser Ausgabe des Skripts wurden die TeX-Sourcen des Dokuments komplett
überarbeitet und natürlich wieder einige Themen korrigiert,
ergänzt und auf den aktuellen Stand gebracht, z.B.:
\vspace{-7pt}
\begin{itemize}
  \item die Suche nach den größten Primzahlen 
        (Kap. \ref{search_for_very_big_primes}),
  \item Fortschritte bei der Kryptoanalyse von AES 
        (Kap. \ref{NeueAES-Analyse}) und
%  \item Fortschritte bei der Kryptoanalyse von Hashverfahren 
%        (Kap. \ref{collision-attacks-against-sha-1}) und
%  \item Fortschritte bei den Ideen für neue Kryptoverfahren (RSA-Nachfolger)
%	(Kap. \ref{xxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxxxxxxx} und
  \item die Auflistung, in welchen Filmen und Romanen Kryptographie eine
        wesentliche Rolle spielt (siehe Anhang \ref{s:appendix-movies}),
        und Kurioses zu Primzahlen (siehe \ref{HT-Quaint-curious-Primes-usage}).
\end{itemize}

@@

% --------------------------------------------------------------------------
\clearpage\phantomsection
\addcontentsline{toc}{chapter}{Einführung -- Zusammenspiel von Skript
                               und CrypTool}
\chapter*{Einführung -- Zusammenspiel von Skript und CrypTool}

### 2009-08-07 script/de/chapters/numbertheory.tex
CT script: Minor updates in the preface (de+en), chap 1 (de+en) and chap 4 (Sage samples, en)

  \item {Hr. Henrik Koy für das anregende und sehr konstruktive 
         Korrekturlesen, für die vielen Verbesserungen dieses Artikels
         und für die Hilfen bei TeX, ohne die dieser Artikel nie in 
         dieser Form erschienen wäre. \\
	 Außerdem hat Hr. Koy in seiner Freizeit im CrypTool-Programm 
	 die komplexe Dialogbox zum RSA-Kryptosystem
%	 RSA-Krypto"-system   (TRENNUNG erzwingen)
	 designed und die dahinterliegende Logik entwickelt, mit der die
	 RSA-Beispiele dieses Artikels nachvollzogen werden können.}

  \item {Jörg Cornelius Schneider für die Unterstützung bei TeX und
         die mannigfaltigen Hilfen bei allen Arten von Programmier- und
	 Design-Problemen.}

  \item {Dr. Georg Illies für den Hinweis auf Pari-GP \index{Pari-GP}.}

>>>>

  \item {Hr. Henrik Koy für das anregende und sehr konstruktive 
         Korrekturlesen und für die vielen Verbesserungen der ersten
	 Version dieses Artikels, und für die Hilfen bei TeX, ohne die
	 dieser Artikel nie in dieser Form erschienen wäre. \\
	 Außerdem hat Hr. Koy in seiner Freizeit im CrypTool-Programm 
	 die komplexe Dialogbox zum RSA-Kryptosystem
%	 RSA-Krypto"-system   (TRENNUNG erzwingen)
	 designed und die dahinterliegende Logik entwickelt, mit der die
	 RSA-Beispiele dieses Artikels nachvollzogen werden können.}

  \item {Jörg Cornelius Schneider für die engagierte Unterstützung bei TeX und
         die mannigfaltigen Hilfen bei allen Arten von Programmier- und
	 Design-Problemen.}

  \item {Dr. Georg Illies für den Hinweis auf Pari-GP\index{Pari-GP}.}

  \item {Lars Fischer für seine Hilfe bei schnellem Pari-GP\index{Pari-GP}-Code
         für Primitivwurzeln.}

  \item {Minh Van Nguyen aus Australien für seine immer schnelle, kompetente und
         ausführliche Hilfe bei den Sage\index{Sage}-Code-Beispielen in diesem
	 Kapitel.}

### 2009-08-07 script/en/chapters/introduction.tex
new float cryptoprocedure for crypto algorithms and procedures (chapter 5); fixed links to un-numbered chapters with phantomsection
CT script: Minor updates in the preface (de+en), chap 1 (de+en) and chap 4 (Sage samples, en)

% --------------------------------------------------------------------------
\chapter*{Preface to the 10th Edition of the CrypTool Script}
\addcontentsline{toc}{chapter}{Preface to the 10th Edition of the CrypTool Script}

@@

%This edition of the script adds some topics, but mainly updates areas (e.g. the
%summaries of topical research areas):
Again this edition of the script corrected, amended and updated some topics, e.g.:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers, 
%  \item progress in cryptanalysis of hash algorithms 
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor) 
%        (chap. \ref{xxxxxxxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory 
        played major role (see appendix \ref{s:appendix-movies});
        and where primes are used as hangers  
        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}).
\end{itemize}

Newly added is the appendix \ref{s:appendix-using-sage} about using the
computer algebra system Sage. This was added because Sage becomes
more and more the standard open-source CAS system. Accordingly all
samples written before in Pari-GP and Mathematica have been substituted
with Sage code. Thanks to Minh Van Nguyen a lot of new code samples could be added.

@@

% --------------------------------------------------------------------------
\newpage
\chapter*{Introduction -- How do the Script and the Program Play together?}
\addcontentsline{toc}{chapter}{Introduction -- How do the Script and the Program Play together?}

>>>>

% --------------------------------------------------------------------------
\clearpage\phantomsection
\addcontentsline{toc}{chapter}{Preface to the 10th Edition of the CrypTool Script}
\chapter*{Preface to the 10th Edition of the CrypTool Script}

@@

%This edition of the script adds some topics, but mainly updates areas (e.g. the
%summaries of topical research areas):
This edition completely updated the TeX sources of the document and of course
the content of the script was corrected, amended and updated with some topics, e.g.:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers 
        (chap. \ref{search_for_very_big_primes}), 
  \item progress in cryptanalysis of hash algorithms 
        (chap. \ref{NeueAES-Analyse}) and
%  \item progress in cryptanalysis of hash algorithms 
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor) 
%        (chap. \ref{xxxxxxxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory 
        played major role (see appendix \ref{s:appendix-movies});
        and where primes are used as hangers  
        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}).
\end{itemize}

Newly added is the appendix \ref{s:appendix-using-sage} about using the
computer algebra system Sage. This was added because Sage becomes
more and more the standard open-source CAS system. Accordingly all
samples written before in Pari-GP and Mathematica have been substituted
with Sage code. Thanks to Minh Van Nguyen, a lot of new code samples could be added.

@@

% --------------------------------------------------------------------------
\clearpage\phantomsection
\addcontentsline{toc}{chapter}{Introduction -- How do the Script and the Program Play together?}
\chapter*{Introduction -- How do the Script and the Program Play together?}

### 2009-08-07 script/en/chapters/numbertheory.tex
CT script: Minor updates in the preface (de+en), chap 1 (de+en) and chap 4 (Sage samples, en)

If you allow numbers other than primes as factors, there are several ways of factorizing integers and the {\em 
uniqueness} is lost:
$$60 = 1*60 = 2*30 = 4*15 = 5*12 = 6*10 = 2*3*10 = 2*5*6 = 3*4*5 = \cdots$$
The 1st fundamental theorem only appears to be obvious. We can construct numerous other sets of 
numbers\footnote{These sets are formed especially from the set of natural numbers. An example of this can be found in this \hyperlink{uniqueness}{script} on 
page~\pageref{thm-pz-euklid} % should be \pageref{remFundTheoOfArithm}, but
% hyperref seems to be buggy
at the end of chapter~\ref{primesinmath}.
}
for which numbers in the 
set {\em cannot} be expressed 
uniquely as a product of the prime numbers of the set.

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: L=enum_PrimitiveRoots_of_an_Integer(541)
{2, 517, 10, 523, 13, 14, 527, 528, 18, 531, 24, 539, 30, 37, 40, 51,
54, 55, 59, 62, 65, 67, 68, 72, 73, 77, 83, 86, 87, 91, 94, 96, 98,
99, 107, 113, 114, 116, 117, 126, 127, 128, 131, 132, 138, 150, 152,
153, 156, 158, 163, 176, 181, 183, 184, 195, 197, 199, 206, 208,
210, 213, 218, 220, 223, 224, 244, 248, 250, 257, 258, 259, 260,
261, 263, 267, 269, 270, 271, 272, 274, 278, 280, 281, 282, 283,
284, 291, 293, 297, 317, 318, 321, 323, 328, 331, 333, 335, 342,
344, 346, 357, 358, 360, 365, 378, 383, 385, 388, 389, 391, 403,
409, 410, 413, 414, 415, 424, 425, 427, 428, 434, 442, 443, 445,
447, 450, 454, 455, 458, 464, 468, 469, 473, 474, 476, 479, 482,
486, 487, 490, 501, 504, 511}
sage: L
sage: len(L)
144
\end{Verbatim}
\caption{Table with all primitive roots for a given prime}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
# BB_Start -- Testcases for count_PrimitiveRoots_of_an_IntegerRange(start, end, bPrimesOnly=True)
print "\n\nBB_Start -- Testcases for count_PrimitiveRoots_of_an_IntegerRange(start, end, bPrimesOnly=True)"

@@

3-----------Testcase: (1, 500, False)
Found all 11010 primitive roots of 170 numbers.
(Total of numbers checked: 500, Amount of numbers without primitive roots: 330)
Time: CPU 1.52 s, Wall: 1.59 s
\end{Verbatim}
\caption{Testcases and output of function ``count\_PrimitiveRoots\_of\_an\_IntegerRange''}
\end{sagecode}

@@

OUTPUT:
CC_Start -- Testcases for count_PrimitiveRoots_of_a_PrimesRange(start, end)
-----------Testcase: (1, 1500)
2 1
3 1
5 2
7 2
11 4
13 4
17 8
19 6
23 10
29 12
31 8
37 12
...
1483 432
1487 742
1489 480
1493 744
1499 636
Found all 62044 primitive roots of 239 primes.
Time: CPU 7.55 s, Wall: 7.85 s
\end{Verbatim}
\caption{Function to calculate all primitive roots for a given range of primes}
\end{sagecode}

\newpage
\newpage
\noindent A slightly modified version of our custom-defined function
\verb!count_PrimitiveRoots_of_a_PrimesRange!, is used to generate
a database of all primitive roots of all primes between 1 and 100,000.

@@

\vskip +10 pt 
\hypertarget{AppArith4a}{%
\noindent \textbf{Example on page~\pageref{SrcArith4a}.}} \\
The RSA exponentiation $M^{37} \pmod{3713}$ on message $M = 120$ can be
calculated in Sage as follows:

@@

\vskip +10 pt 
\hypertarget{AppArith4b}{%
\noindent {\bf Example on page~\pageref{SrcArith4b}.}} \\
The factorization of $J(256027) = 255016 = 2^3 * 127 * 251$ can be
calculated using Sage as follows:

@@

\vskip +10 pt 
\hypertarget{AppArith4c}{%
\noindent {\bf Example on page~\pageref{SrcArith4c}.}} \\
Sage can do RSA encryption as follows:

@@

\vskip +10 pt 
\hypertarget{AppArith4d}{%
\noindent {\bf Example on page~\pageref{SrcArith4d}.}} \\
RSA encryption using Sage:

@@

\vskip +10 pt 
\hypertarget{AppArith4e}{%
\noindent {\bf Example on page~\pageref{SrcArith4e}.}} \\
RSA encryption using Sage:

@@

         \href{http://www.rsa.com/rsalabs/node.asp?id=2092}
           {\texttt{http://www.rsa.com/rsalabs/node.asp?id=2092}}

@@

@@

  \item {Henrik Koy for making many very useful suggestions, 
         for the very constructive proof-reading of this article and
         for helping with TeX. \\
	 Mr.\ Koy designed and developed in his leisure time the 
	 functions and the complex dialog box of the RSA cryptosystem within CrypTool, 
	 which enables you to execute the RSA samples of this article.}

@@

  \item {Minh Van Nguyen from Australia for his always fast and exhaustive
         help with the Sage\index{Sage} code samples in this chapter.}

>>>>

If you allow numbers other than primes as factors, there are several ways of factorizing integers and the {\em 
uniqueness} is lost:
$$60 = 1*60 = 2*30 = 4*15 = 5*12 = 6*10 = 2*3*10 = 2*5*6 = 3*4*5 = \cdots$$
The 1st fundamental theorem only appears to be obvious. We can construct numerous other sets of 
numbers\footnote{These sets are formed especially from the set of natural numbers. An example of
this can be found in this \hyperlink{uniqueness}{script} on 
page~\pageref{thm-pz-euklid} % TODO: should be \pageref{remFundTheoOfArithm}, but
                             % hyperref seems to be buggy
at the end of chapter~\ref{primesinmath}.
}
for which numbers in the 
set {\em cannot} be expressed 
uniquely as a product of the prime numbers of the set.

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: L=enum_PrimitiveRoots_of_an_Integer(541)
{2, 517, 10, 523, 13, 14, 527, 528, 18, 531, 24, 539, 30, 37, 40, 51,
54, 55, 59, 62, 65, 67, 68, 72, 73, 77, 83, 86, 87, 91, 94, 96, 98,
99, 107, 113, 114, 116, 117, 126, 127, 128, 131, 132, 138, 150, 152,
153, 156, 158, 163, 176, 181, 183, 184, 195, 197, 199, 206, 208,
210, 213, 218, 220, 223, 224, 244, 248, 250, 257, 258, 259, 260,
261, 263, 267, 269, 270, 271, 272, 274, 278, 280, 281, 282, 283,
284, 291, 293, 297, 317, 318, 321, 323, 328, 331, 333, 335, 342,
344, 346, 357, 358, 360, 365, 378, 383, 385, 388, 389, 391, 403,
409, 410, 413, 414, 415, 424, 425, 427, 428, 434, 442, 443, 445,
447, 450, 454, 455, 458, 464, 468, 469, 473, 474, 476, 479, 482,
486, 487, 490, 501, 504, 511}
sage: L
sage: len(L)
144
\end{Verbatim}
\caption{Table with all primitive roots for the given prime 541}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
# BB_Start -- Testcases for count_PrimitiveRoots_of_an_IntegerRange(start, end, bPrimesOnly=True)
print "\n\nBB_Start -- Testcases for count_PrimitiveRoots_of_an_IntegerRange(start, end, True)"

@@

3-----------Testcase: (1, 500, False)
Found all 11010 primitive roots of 170 numbers.
(Total of numbers checked: 500, Amount of numbers without primitive roots: 330)
Time: CPU 1.52 s, Wall: 1.59 s
\end{Verbatim}
\caption{Function ``count\_PrimitiveRoots\_of\_an\_IntegerRange'': testcases and output}
\end{sagecode}

@@

OUTPUT:
CC_Start -- Testcases for count_PrimitiveRoots_of_a_PrimesRange(start, end)
-----------Testcase: (1, 1500)
2 1
3 1
5 2
7 2
11 4
13 4
17 8
19 6
23 10
29 12
31 8
37 12
...
1483 432
1487 742
1489 480
1493 744
1499 636
Found all 62044 primitive roots of 239 primes.
Time: CPU 7.55 s, Wall: 7.85 s
\end{Verbatim}
\caption{Function ``count\_PrimitiveRoots\_of\_a\_PrimesRange'' to calculate all primitive roots for a given range of primes}
\end{sagecode}

\newpage
\newpage
\noindent A slightly modified version of our function
\verb!count_PrimitiveRoots_of_a_PrimesRange!, is used to generate
a database of all primitive roots of all primes between 1 and 100,000.

@@

\vskip +10 pt 
\hypertarget{AppArith4a}{%
\noindent \textbf{Example on page~\pageref{SrcArith4a}:}} \\
The RSA exponentiation $M^{37} \pmod{3713}$ on message $M = 120$ can be
calculated in Sage as follows:

@@

\vskip +10 pt 
\hypertarget{AppArith4b}{%
\noindent {\bf Example on page~\pageref{SrcArith4b}:}} \\
The factorization of $J(256027) = 255016 = 2^3 * 127 * 251$ can be
calculated using Sage as follows:

@@

\vskip +10 pt 
\hypertarget{AppArith4c}{%
\noindent {\bf Example on page~\pageref{SrcArith4c}:}} \\
Sage can do RSA encryption as follows:

@@

\vskip +10 pt 
\hypertarget{AppArith4d}{%
\noindent {\bf Example on page~\pageref{SrcArith4d}:}} \\
RSA encryption using Sage:

@@

\vskip +10 pt 
\hypertarget{AppArith4e}{%
\noindent {\bf Example on page~\pageref{SrcArith4e}:}} \\
RSA encryption using Sage:

@@

         \href{http://www.rsa.com/rsalabs/node.asp?id=2092}
           {\texttt{http://www.rsa.com/rsalabs/node.asp?id=2092}}\\

@@

   \item Sage, \index{Sage} \\
         Excellent, open source computer algebra system with Python as script language,
	 used to build the code samples in this chapter.
	 See also chapter~\ref{s:appendix-using-sage}.\\
         \href{http://www.sagemath.org/}
              {\texttt{http://www.sagemath.org/}}\\
         \href{http://en.wikipedia.org/wiki/Sage_%28mathematics_software%29}
              {\texttt{http://en.wikipedia.org/wiki/Sage\_\%28mathematics\_software\%29}}\\

@@

  \item {Henrik Koy for making many very useful suggestions, 
         for the very constructive proof-reading of the first version of
	 this article, and for helping with TeX. \\
	 Mr.\ Koy designed and developed in his leisure time the 
	 functions and the complex dialog box of the RSA cryptosystem within CrypTool, 
	 which enables you to execute the RSA samples of this article.}

@@

  \item {Minh Van Nguyen from Australia for his always fast, professional and exhaustive
         help with the Sage\index{Sage} code samples in this chapter.}
